---
editor_options:
  chunk_output_type: inline
output:
  pdf_document: 
    latex_engine: lualatex
  html_document: 
    toc: yes
---

# Homework 3

*Ginevra Carbone*

# LEC

## Exercise 1

Compute the bootstrap-based confidence interval for the `score` dataset using the studentized method.
```{r echo=TRUE}
score <- read.table("student_score.txt", header = TRUE)
psi_fun <- function(data) {
  eig <- eigen(cor(data))$values
  return(max(eig) / sum(eig))
}

psi_obs <- psi_fun(score)

# bootstrap
n <- nrow(score); B <- 10^4
s_vect <- rep(0, B)
for(i in 1:B) {
  ind <- sample(1:n, n, replace = TRUE)
  s_vect[i] <- psi_fun(score[ind,])
}
SE_boot <- sd(s_vect)

# bootstrap for z_vect
n <- nrow(score)
new_s_vect <- rep(0, n)
for(i in 1:B) {
  ind <- sample(1:n, n, replace = TRUE)
  new_s_vect[i] <- psi_fun(score[ind,])
}
SE_z_vect <- sd(new_s_vect)
z_vect <- (s_vect - psi_obs)/SE_z_vect

student_ci <- psi_obs - SE_boot * quantile(z_vect, prob=c(0.975, 0.025))
student_ci
```

## Exercise 2

Compute bootstrap-based confidence intervals for the `score` dataset using the `boot` package.
```{r echo=TRUE}
library(boot)

score <- read.table("student_score.txt", header = TRUE)

psi_fun <- function(data, ind) {
  d <- data[ind,] # sample selection
  eig <- eigen(cor(d))$values
  return(max(eig) / sum(eig))
}

# bootstrapping with 1000 replications 
results <- boot(data = score, statistic = psi_fun, R = 1000)

# view results
results
plot(results)

# get 95% confidence interval 
boot.ci(results, type = c("perc","basic"))

psi_fun <- function(data, ind) {
  d <- data[ind,] # sample selection
  eig <- eigen(cor(d))$values
  out <- max(eig)/sum(eig)
  # return(c(out, var(out)))
  return(out)
}

psi_fun_var <- function(data, ind){
  d <- data[ind,] 
  boot <- boot(data = score, statistic = psi_fun, R = 50)
  eig <- eigen(cor(d))$values
  out <- max(eig)/sum(eig)
  return(c(out, var(boot$t)))
}

resultsVar <- boot(data = score, statistic = psi_fun_var, R = 100)

boot.ci(resultsVar, type = "stud")
```

# LAB 

## Exercise 1

Compute analitically $J(\gamma,\gamma;\gamma), J(\gamma,\beta;y), J(\beta,\beta;y)$.

$$
J(\theta;y) = 
\begin{pmatrix}
\frac{n}{\gamma^2}+\sum_{i=1}^n \big(\frac{y_i}{\beta}\big)^\gamma \big( log \frac{y_i}{\beta} \big)^2 && \frac{n}{\beta}-\sum_{i=1}^n \frac{y_i ^\gamma}{\beta^{\gamma+1}}\big(1+ \gamma\, log \frac{y_i}{\beta}\big) \\
\frac{n}{\beta}-\sum_{i=1}^n \frac{y_i ^\gamma}{\beta^{\gamma+1}}\big(1+ \gamma\, log \frac{y_i}{\beta}\big) && -\frac{n\gamma}{\beta^2}+\gamma \frac{\gamma+1}{\beta^{\gamma+2}}\sum_{i=1}^n y_i ^\gamma
\end{pmatrix}
$$

## Exercise 2

Produce the contour plot for the quadratic approximation of the log-likelihood, based on the Taylor series:

$$ l(\theta)-l(\hat{\theta}) \approx -\frac{1}{2} (\theta-\hat{\theta})^T J(\hat{\theta}) (\theta-\hat{\theta}) $$

```{r echo=TRUE}
# observations
y <- c(155.9, 200.2, 143.8, 150.1,152.1, 142.2, 147, 146, 146,
170.3, 148, 140, 118, 144, 97)
n <- length(y)

# thetahat
gammahat<-uniroot(function(x) n/x+sum(log(y))-n* sum(y^x*log(y))/sum(y^x), c(1e-5,15))$root 
betahat<- mean(y^gammahat)^(1/gammahat)
weib.y.mle<-c(gammahat,betahat) # thetahat

# observed information matrix
jhat<-matrix(NA,nrow=2,ncol=2)
jhat[1,1]<-n/gammahat^2+sum((y/betahat)^gammahat* (log(y/betahat))^2)
jhat[1,2]<-jhat[2,1]<- n/betahat-sum(y^gammahat/betahat^(gammahat+1)*
(gammahat*log(y/betahat)+1))
jhat[2,2]<- -n*gammahat/betahat^2+gammahat*(gammahat+1)/
betahat^(gammahat+2)*sum(y^gammahat)
jhat <- as.matrix(jhat)

# parameters grid
gamma <- seq(0.1, 15, length=100)
beta <- seq(100, 200, length=100)
parvalues <- expand.grid(gamma,beta)

# quadratic approximation function
approximation <- function(theta){
diff <- theta - weib.y.mle
return(-0.5*t(diff)%*%jhat%*%diff)
}

quadratic_approx <- apply(parvalues, 1, approximation)
quadratic_approx <- matrix(quadratic_approx,nrow=length(gamma), ncol=length(beta), byrow=F )

#contour plot
conf.levels <- c(0,0.5,0.75,0.9,0.95,0.99)
contour(gamma, beta, quadratic_approx,
        levels=-qchisq(conf.levels, 2)/2,
        xlab=expression(gamma),
        labels=as.character(conf.levels),
        ylab=expression(beta))
title('Weibull relative log likelihood')

```

## Exercise 3 

Use `nlm` for computing the variance for the estimator $\hat{w} = (log(\hat{\gamma}), log(\hat{\beta}))$ and `optimHess` for the variance of $\hat{\theta} = (\hat{\gamma},\hat{\beta})$.

```{r echo=TRUE, warning=FALSE}
y <- c(155.9, 200.2, 143.8, 150.1,152.1, 142.2, 147, 146, 146,
170.3, 148, 140, 118, 144, 97)
n <- length(y)

log_lik_weibull <- function(data, param){
  -sum(dweibull(data, shape = param[1], scale = param[2], log = TRUE))
}

omega <- function(theta) log(theta)
theta <- function(omega) exp(omega)
log_lik_weibull_rep <- function(data, param) log_lik_weibull(data, theta(param))

weib.y.nlm<-nlm(log_lik_weibull_rep, c(0,0), hessian=T,data=y)

# variance of omegahat
hess <- weib.y.nlm$hessian
diag(solve(hess))

# variance of thetahat
hess <- optimHess(theta(weib.y.nlm$estimate), log_lik_weibull, data=y)
diag(solve(hess))
```

## Exercise 4

The Wald confidence interval with level $1-\alpha$ is defined as:
$$ \hat{\gamma}\pm z_{1-\alpha/2}\,j_P(\hat{\gamma})^{-1/2} $$
Compute the Wald confidence interval of level 0.95 and plot the results.

```{r echo=TRUE}
weib.y.mle<-optim(c(1,1), fn=log_lik_weibull,hessian=T, method='L-BFGS-B',lower=rep(1e-7,2), upper=rep(Inf,2),data=y) 

conf.level<-0.95
# SE for the estimate from the observed information matrix
weib.y.se <- sqrt(diag(solve(weib.y.mle$hessian)))
# Wald confidence interval for gamma
wald.ci1 <- weib.y.mle$par[1]+c(-1,1)*qnorm(1-(1-conf.level)/2)*weib.y.se[1]
wald.ci1

# plot
log_lik_weibull_profile  <- function(data, gamma){
  beta.gamma <- mean(data^gamma)^(1/gamma)
  log_lik_weibull( data, c(gamma, beta.gamma) )
}

log_lik_weibull_profile_v <-Vectorize(log_lik_weibull_profile, 'gamma'  )

plot(function(x) 
  -log_lik_weibull_profile_v(data=y, x)
  +weib.y.mle$value, 
  from=0.1,to=15,xlab=expression(gamma),
  ylab='profile relative log likelihood',ylim=c(-8,0))

segments( wald.ci1[1],
          -log_lik_weibull_profile_v(y,wald.ci1[1])
          -weib.y.mle$value, wald.ci1[1], 
          -log_lik_weibull_profile_v(y,wald.ci1[1])
          +weib.y.mle$value, col="blue", lty=2  )

segments( wald.ci1[2],
          -log_lik_weibull_profile_v(y,wald.ci1[2])
          -weib.y.mle$value, wald.ci1[2], 
          -log_lik_weibull_profile_v(y, wald.ci1[2])
          +weib.y.mle$value, col="blue", lty=2  )

segments( wald.ci1[1], -8,  wald.ci1[2], -8, col="blue", lty =1, lwd=2  )
text(7,-7.5,"95% Wald CI",col="blue")

```


## Exercise 5

Repeat the steps above — write the profile log-likelihood, plot it and find the deviance confidence intervals — considering this time $\gamma$ as a nuisance parameter and $\beta$ as the parameter of interest.

```{r echo=TRUE}
weib.y.mle<-optim(c(1,1), fn=log_lik_weibull,hessian=T, method='L-BFGS-B',lower=rep(1e-7,2), upper=rep(Inf,2),data=y) 

log_lik_weibull_profile  <- function(data, beta){
  # gamma is fixed
  gamma.beta <- uniroot(function(x) n/x+sum(log(data))-n* sum(data^x*log(data))/sum(data^x),c(1e-5,15))$root
  log_lik_weibull( data, c(gamma.beta, beta) )
}

log_lik_weibull_profile_v <- Vectorize(log_lik_weibull_profile, 'beta' )

plot(function(x) -log_lik_weibull_profile_v(data=y, x)
     +weib.y.mle$value,
     from=120,to=200,xlab=expression(beta),
     ylab='profile relative log likelihood',
     ylim=c(-10,0))

conf.level<-0.95
abline(h=-qchisq(conf.level,1)/2,lty='dashed',col=2)

lrt.ci1 <- uniroot(function(x) 
  -log_lik_weibull_profile_v(y, x)
  + weib.y.mle$value
  + qchisq(conf.level,1)/2,
  c(1e+7,weib.y.mle$par[2]))$root

lrt.ci1<-c(lrt.ci1,uniroot(function(x)
  -log_lik_weibull_profile_v(y,x)
  + weib.y.mle$value
  + qchisq(conf.level,1)/2,
  c(weib.y.mle$par[2],15))$root)

segments( lrt.ci1[1],-qchisq(conf.level,1)/2, lrt.ci1[1], -log_lik_weibull_profile_v(y, lrt.ci1[1]), col="red", lty=2  )

segments( lrt.ci1[2],-qchisq(conf.level,1)/2, lrt.ci1[2], -log_lik_weibull_profile_v(y, lrt.ci1[2]), col="red", lty=2  )

points(lrt.ci1[1], -qchisq(0.95,1)/2, pch=16, col=2, cex=1.5)
points(lrt.ci1[2], -qchisq(0.95,1)/2, pch=16, col=2, cex=1.5)
segments( lrt.ci1[1], -10.1, lrt.ci1[2], -10.1, col="red", lty =1, lwd=2  )
text(157,-9.5,"95% Deviance CI",col=2, cex=0.8)

```

## Exercise 6

Perform a test as above, but with:
$$\begin{cases}
H_0 : \gamma = 1 \\
H_1 : \gamma = 5
\end{cases}$$

```{r echo=TRUE}
lambda_lrt <- -2*(log_lik_weibull_profile(y,5)-log_lik_weibull_profile(y,1))
lambda_lrt

p_lrt <- pchisq(lambda_lrt, df =1, lower.tail = FALSE)
p_lrt
# We end up rejecting the null hypotesis, since gamma is closer to 5
```
